---
title: "Hwk2_nikhila_ravi"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

### Load the data:

```{r cars}
taxi_train <- read.table('dataset_1_train.txt', sep = ',', header =TRUE)
taxi_test <- read.table('dataset_1_test.txt', sep = ',', header =TRUE)
taxi_train
```

### Visualise the data:

```{r cars}
library(ggplot2)
ggplot(taxi_train, aes(x = DayOfWeek, y = PickupCount)) + 
  geom_point()
ggplot(taxi_train, aes(x = TimeMin, y = PickupCount)) + 
  geom_point()
```

## Question 1a

### Helper functions

```{r}
rsq = function(model, data, y) {
  y <- data[[y]]
  predict <- predict(model, newdata = data)
  tss = sum((y - mean(y))^2)
  rss = sum((y-predict)^2)
  rsq_ = max(0, 1 - rss/tss)
  return(rsq_)
}
```

### Regression models: polynomials

```{r}
poly_3 <- lm(PickupCount ~ poly(TimeMin, degree = 3, raw = TRUE), data = taxi_train)
poly_10 <- lm(PickupCount ~ poly(TimeMin, degree = 10, raw = TRUE), data = taxi_train)
poly_25 <- lm(PickupCount ~ poly(TimeMin, degree = 25, raw = TRUE), data = taxi_train)

taxi.predict <- data.frame(taxi_train, pred_3 = predict(poly_3), pred_10 = predict(poly_10), pred_25 = predict(poly_25))
ggplot(taxi.predict, aes(x = TimeMin, y=PickupCount)) + 
  geom_point() + 
  geom_line(aes(y = pred_3), colour="blue") + 
  geom_line(aes(y = pred_10), colour = "green") + 
  geom_line(aes(y = pred_25), colour = "red") + 
  ylab(label="PickupCount") + 
  xlab("TimeMin")
```

#### Analyse the R2 of the models 

```{r}
accuracy <- data.frame('model'='3rd degree poly', 'train_rsquared'=rsq(poly_3, taxi_train,'PickupCount'), 'test_rsquared'=rsq(poly_3, taxi_test,'PickupCount'))
accuracy <- rbind(accuracy, data.frame('model' = '10th degree poly', 'train_rsquared'=rsq(poly_10, taxi_train,'PickupCount'), 'test_rsquared'=rsq(poly_10, taxi_test,'PickupCount')))
accuracy <- rbind(accuracy, data.frame('model' = '25th degree poly', 'train_rsquared'=rsq(poly_25, taxi_train,'PickupCount'), 'test_rsquared'=rsq(poly_25, taxi_test,'PickupCount')))
```

### Regression models: cubic b splines

```{r}
library(splines)
quarts = quantile(taxi_train$TimeMin,probs=c(0.4,0.5,0.7))
cubicbspline=lm(PickupCount~bs(TimeMin,knots=quarts),data=taxi_train)

accuracy <- rbind(accuracy, data.frame('model' = 'Cubic spline', 'train_rsquared'=rsq(cubicbspline, taxi_train,'PickupCount'), 'test_rsquared'=rsq(cubicbspline, taxi_test,'PickupCount')))


taxi.predict['cubicsplinepred'] <- predict(cubicbspline)
ggplot(taxi.predict, aes(x = TimeMin, y=PickupCount)) + 
  geom_point() + 
  geom_line(aes(y = cubicsplinepred), colour="blue") + 
  ylab(label="PickupCount") + 
  xlab("TimeMin")
```

```{r}
accuracy
```

### Regression models: natural cubic splines

```{r}
## degrees of freedon

dfs <- 1:6
## create k partitions
taxi_train$splits <- cut(sample(1:nrow(taxi_train), 
                            nrow(taxi_train)), 
                     breaks = 5, 
                     labels = FALSE)

model.performance <- function(df, train, test) {
    mod <- lm(PickupCount ~ ns(TimeMin, df = df), data = train)
    c(train.r2 = rsq(mod, train, "PickupCount"),
      test.r2 = rsq(mod, test, "PickupCount"))
}

## iterate over the splits, holding each one out as the test set.
perform.5fold <- lapply(
  unique(taxi_train$splits), 
  function(split) {
    train <- taxi_train$split == split
    mt.train <- taxi_train[train, ]
    mt.test <- taxi_train[-train, ]
    data.frame(t(sapply(dfs, model.performance, 
                   train = mt.train,
                   test = mt.test)),
          df = dfs)
  }
)

## collect the k sets of model statistics in a data.frame
perform.5fold <- do.call(rbind, perform.5fold)

## aggregate across the k sets, averaging model statistics for each df
perform.5fold <- lapply(split(perform.5fold, perform.5fold$df),
                        function(x) {
                          data.frame(rsquare = c(mean(x$train.r2),
                                                 mean(x$test.r2)),
                                     data = c("train", "test"),
                                     df = unique(x$df))
                        })

## collect the results
perform.5fold <- do.call(rbind, perform.5fold)

## plot the results
ggplot(perform.5fold, aes(x = df, y = rsquare, color = data)) +
  geom_point()
```

```{r}
nspline <- lm(PickupCount ~ ns(TimeMin, df = 6), data = taxi_train)
accuracy <- rbind(accuracy, data.frame('model' = 'Natural Cubic spline with 6 degree of freedom', 'train_rsquared'=rsq(nspline, taxi_train,'PickupCount'), 'test_rsquared'=rsq(nspline, taxi_test,'PickupCount')))
accuracy
```


### Smoothing Splines

```{r}
spline = smooth.spline(taxi_train$TimeMin, taxi_train$PickupCount, cv=TRUE)
spline['spar']
```
```{r}
fitsmoothspline = smooth.spline(taxi_train$TimeMin, taxi_train$PickupCount, spar = spline['spar'])

splinersq = function(pred, y) {
  tss = sum((y - mean(y))^2)
  rss = sum((y-pred)^2)
  rsq_ = max(0, 1 - rss/tss)
  return(rsq_)
}
pred_train <- predict(fitsmoothspline, taxi_train$TimeMin)$y
pred_test <- predict(fitsmoothspline, taxi_test$TimeMin)$y

accuracy <- rbind(accuracy, data.frame('model' = 'Smoothing spline', 'train_rsquared'=splinersq(pred_train, taxi_train$PickupCount), 'test_rsquared'=splinersq(pred_test, taxi_test$PickupCount)))
```

```{r}
accuracy
```


### Locally Weighted regression model

```{r}
loess_5foldcv = function(data_set){
  data_set$splits <- cut(sample(1:nrow(data_set), 
                            nrow(data_set)), 
                     breaks = 5, 
                     labels = FALSE)
  spans <- seq(0.1, 0.8, by = 0.1)

## iterate over the splits, holding each one out as the test set.
perform.5fold <- list()
for(split in data_set$splits) {
  train <- data_set$split == split
  mt.train <- data_set[train, ]
  mt.test <- data_set[-train, ]
  
  train.r2 <- c()
  test.r2 <- c()
  
  for(sp in spans) {
    mod <- loess(PickupCount ~ TimeMin, span = sp, data = mt.train)
    train.r2 <- c(train.r2, rsq(mod, mt.train, "PickupCount"))
    test.r2 <- c(test.r2, rsq(mod, mt.test, "PickupCount"))
  }
  
  loess.r2 <- data.frame(span = c(spans, spans),
                       set = rep(c("train", "test"), each = length(spans)),
                       r2 = c(train.r2, test.r2))
  perform.5fold <- c(perform.5fold, list(loess.r2))
}

## collect the k sets of model statistics in a data.frame
perform.5fold <- do.call(rbind, perform.5fold)

## aggregate across the k sets, averaging model statistics for each df
perform.5fold <- aggregate(perform.5fold$r2, perform.5fold[c("span", "set")], FUN = mean)

## plot the results
ggplot(perform.5fold, aes(x = span, y = x)) +
  geom_point() +
  geom_line() +
  ylab(expression(R^2))
}
```


```{r}
spans <- seq(0.1, 0.8, by = 0.1)

## iterate over the splits, holding each one out as the test set.
perform.5fold <- list()
for(split in taxi_train$splits) {
  train <- taxi_train$split == split
  mt.train <- taxi_train[train, ]
  mt.test <- taxi_train[-train, ]
  
  train.r2 <- c()
  test.r2 <- c()
  
  for(sp in spans) {
    mod <- loess(PickupCount ~ TimeMin, span = sp, data = mt.train)
    train.r2 <- c(train.r2, rsq(mod, mt.train, "PickupCount"))
    test.r2 <- c(test.r2, rsq(mod, mt.test, "PickupCount"))
  }
  
  loess.r2 <- data.frame(span = c(spans, spans),
                       set = rep(c("train", "test"), each = length(spans)),
                       r2 = c(train.r2, test.r2))
  perform.5fold <- c(perform.5fold, list(loess.r2))
}

## collect the k sets of model statistics in a data.frame
perform.5fold <- do.call(rbind, perform.5fold)

## aggregate across the k sets, averaging model statistics for each df
perform.5fold <- aggregate(perform.5fold$r2, perform.5fold[c("span", "set")], FUN = mean)

## plot the results
ggplot(perform.5fold, aes(x = span, y = x, color = set)) +
  geom_point() +
  geom_line() +
  ylab(expression(R^2))
```

A span of 0.1 has the highest R^2 value. 

```{r}
loess_model <- loess(PickupCount ~ TimeMin, span = 0.1, data = taxi_train)
accuracy <- rbind(accuracy, data.frame('model' = 'Locally weighted regression model', 'train_rsquared'=rsq(loess_model, taxi_train,'PickupCount'), 'test_rsquared'=rsq(loess_model, taxi_test,'PickupCount')))
accuracy
```

```{r}
rsq(loess_model, taxi_train,'PickupCount')
```

### Part 1b: Adapting to Weekends

The time series plots of taxi pickups for each day of the week are shown below:

```{r}
for (day in 1:7){
  df = taxi_train[taxi_train$DayOfWeek == day, ]
  print(ggplot(df, aes(x = TimeMin, y = PickupCount)) +
  geom_point() +
    ggtitle('day of week ', day))
}
```

The time series of weekday vs weekend is also shown below:

```{r}
weekday_train_df = taxi_train[taxi_train$DayOfWeek != 6 & taxi_train$DayOfWeek != 7 , ]
weekend_train_df = taxi_train[taxi_train$DayOfWeek == 6 | taxi_train$DayOfWeek == 7, ]
weekday_test_df = taxi_test[taxi_test$DayOfWeek != 6 & taxi_test$DayOfWeek != 7 , ]
weekend_test_df = taxi_test[taxi_test$DayOfWeek == 6 | taxi_test$DayOfWeek == 7, ]
```

```{r}
ggplot(weekday_train_df, aes(x = TimeMin, y = PickupCount)) +
  geom_point() +
    ggtitle('Weekday pickups ', day) + ylim(0, 100)
ggplot(weekend_train_df, aes(x = TimeMin, y = PickupCount)) +
  geom_point() +
    ggtitle('Weekend pickups', day) + ylim(0, 100)
```

The pattern of pickups is different on the weekends compared to the weekdays

#### Fit locally weighted least squares to each subset

Weekday model

```{r}
loess_5foldcv(weekday_train_df)
```

```{r}
loess_5foldcv(weekend_train_df)
```

```{r}
weekend_model = loess(PickupCount ~ TimeMin, span = 0.1, data = weekend_train_df)
weekday_model = loess(PickupCount ~ TimeMin, span = 0.1, data = weekday_train_df)
```

#### Comparison of R2 values of the new models on the separated test sets compared to model fit on the entire test set

```{r}
loess_accuracy <- data.frame('model'='loess', 'test_rsquared'=rsq(loess_model, taxi_test,'PickupCount'))
loess_accuracy <- rbind(loess_accuracy, data.frame('model'='Loess weekday model on weekday test set', 'test_rsquared'=rsq(weekday_model, weekday_test_df,'PickupCount')))
loess_accuracy <- rbind(loess_accuracy, data.frame('model'='Loess weekend model on weekend test set', 'test_rsquared'=rsq(weekend_model, weekend_test_df,'PickupCount')))
loess_accuracy <- rbind(loess_accuracy, data.frame('model'='Loess model on weekend test set', 'test_rsquared'=rsq(loess_model, weekend_test_df,'PickupCount')))
loess_accuracy <- rbind(loess_accuracy, data.frame('model'='Loess model on weekday test set', 'test_rsquared'=rsq(loess_model, weekday_test_df,'PickupCount')))
```

```{r}
loess_accuracy
```

### Problem 2: Predicting Crime in the City



